Supporting Information

### 2015-08-19 manuscript.Rmd
New SI figure on detectable difference.

## Response of LAI to elevated [CO~2~]
We report on `r nummon` months of continuous leaf area index (LAI) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~), including 3 months of ramp-up in C~a~ from November 2012 to January 2013. We predicted that LAI would increase in response to eC~a~ as a result of enhanced CO~2~ assimilation for the entire canopy and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low LAI [@Whitehead2004], would be even more responsive to eC~a~ than higher-LAI and more closed-canopy stands [@Norby2011]. Counter to our expectations but similar to Norby et al. [-@Norby2003], we found no significant effect of eC~a~ on LAI. Because estimated LL averaged `r round(mean(ba$LL),1)` yrs (Fig. 5), the canopy has regrown at least twice during the study period, giving ample opportunity for eC~a~ effects to occur. It could be argued that the change in LAI in response to eC~a~ was too small to be detected. Natural variation in LAI among plots in native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used pre-treatment stand basal area as a covariate. Using this method, we estimated that we can detect a difference of only a `r format(Ldetect,digits=1)`% in LAI. Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to eC~a~.

** rethink the following; CI on least-square means accounting for BA?**
But what difference between C~a~ treatments could we have reasonably detected? The average half-width of the confidence interval on the LAI anomaly (Fig. 4) was `r f(LanomCIamb)` m^2^m^-2^ for aC~a~, and `r f(LanomCIele)` m^2^m^-2^ for the eC~a~ treatment. This means that, provided the within-treatment variance stays the same, we could detect a difference of `r f(LanomCIamb + LanomCIele)` m^2^ m^-2^ (the sum of the half-width CI's), or about a change of `r f(Ldetect)`% over the mean LAI of the EucFACE (`r f(mean(ba$LAI.mean))` m^2^m^-2^).

@@

**Figure 2**. Time series of LAI estimated from canopy transmittance. Each symbol is a value for a FACE ring, but the smooth curve (fitted with a generalized additive model) takes into account the random plot effect. Note that the 95% confidence always overlap, denoting no significant treatment effect at $\alpha$ = 0.05.

@@

>>>>

## Response of LAI to elevated [CO~2~]
We report on `r nummon` months of continuous leaf area index (LAI) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~), including 3 months of ramp-up in C~a~ from November 2012 to January 2013. We predicted that LAI would increase in response to eC~a~ as a result of enhanced CO~2~ assimilation for the entire canopy and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low LAI [@Whitehead2004], would be even more responsive to eC~a~ than higher-LAI and more closed-canopy stands [@Norby2011]. Counter to our expectations but similar to Norby et al. [-@Norby2003], we found no significant effect of eC~a~ on LAI. Because estimated LL averaged `r round(mean(ba$LL),1)` yrs (Fig. 5), the canopy has regrown at least twice during the study period, giving ample opportunity for eC~a~ effects to occur. It could be argued that the change in LAI in response to eC~a~ was too small to be detected. Natural variation in LAI among plots in native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used pre-treatment stand basal area as a covariate. Using this method, we estimated that we can detect a difference of only a 6.1% in LAI (see Fig. S5). Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to eC~a~.

@@

**Figure 2**. Time series of LAI estimated from canopy transmittance. Each symbol is a value for a FACE ring, but the smooth curve (fitted with a generalized additive model) takes into account the random plot effect. Note that the 95% confidence always overlap, denoting no significant treatment effect at $\alpha$ = 0.05. Variation between rings was substantial, and largely due to pre-treatment differences in stand basal area (see Fig. S3 and S5).

@@

**Figure SI5**. Detectable differences in LAI between C~a~ treatments after accounting for pre-treatment variation in basal area.

### 2015-08-19 manuscript_supportinginfo.Rmd
New SI figure on detectable difference.

```{r echo=FALSE, warning=FALSE, message=FALSE}
source("manuscript_calcs.R")
```

@@

>>>>

```{r echo=FALSE, warning=FALSE, message=FALSE}
source("manuscript_calcs.R")
opts_chunk$set(dpi=300)
```

@@

\
```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE}
res <- figureSI5(df, ba)
```

**Figure SI5**. Detectable differences in LAI between C~a~ treatments after accounting for pre-treatment variation in basal area. We fit LAI with a fitted GAM using C~a~, time and pre-treatment basal area (BA) as the covariates. In the GAM, time was entered as a smooth term (estimated by a cubic spline), and BA as a linear term. Shown are the predictions (smooth lines) from the fitted GAM when BA was set to the mean value across all plots. Shaded areas indicate 95% confidence intervals (CI). Note they are much narrower than the CI across plots when BA was allowed to vary (Fig. 2 in main text). The CI here thus accounts for the pre-treatment differences between rings, and visualizes how using BA as a covariate in the mixed-effects model increases the ability to resolve small difference in LAI that may develop over time. The minimum detectable difference in LAI between C~a~ treatments can be estimated from the average half-width of the CI of fitted LAI at a common BA. A significant effect would could be observed (provided the error variance remains the same) when the two CIs (for aC~a~ and eC~a~) would just not overlap (so that p < 0.05), or twice the half-width CI. This difference was estimated as `r round(res$halfci,2)` m^2^ m^-2^, which when expressed relative to the mean LAI of `r round(mean(ba$LAI.mean),2)` is a difference of only `r round(100*res$minobsdiff,1)`%.
